Persistence in functional topology and a correction to a theorem of Morse

### 2021-07-15 persistence.tex
Improve sect 2

By definition, applying to a filtered space $\{X_t\}_{t \in \R}$ a functor from $\Top$ to $\Vect$ yields for every $t \in \R$ a vector space~$M_t$ and for any pair~$s, t \in \R$ with~$s \leq t$ a linear map~$M_{s,t} \colon M_s \to M_t$ such that $M_{t,t}$ is the identity and the composition $M_{s,t} \circ M_{r,s}$ is equal to $M_{r,t}$ for any triple $r \leq s \leq t$.
In other words, we obtain a functor from the real numbers, considered as a poset category, to the category of vector spaces.
Such functors will be called \emph{persistence modules}.
A morphism of persistence modules $\varphi \colon M \to N$ is a natural transformation, i.e., an assignment of a linear map $\varphi_t \colon M_t \to N_t$ for every $t \in \R$ making the diagram
\begin{equation*}
    \begin{tikzcd}
    M_{s} \arrow[r, "M_{s,t}"] \arrow[d] & M_{t} \arrow[d] \\
    N_{s} \arrow[r, "N_{s,t}"] & N_{t}
    \end{tikzcd}
\end{equation*}
commute for all pairs $s \leq t$.

Usual constructions that work in the category of vector spaces over the field $\F$ can be transferred to the category of persistence modules by applying them pointwise.
For example, the kernel and cokernel of a morphism, as well as the direct sum of persistence modules are well-defined.
Persistence modules that are \emph{indecomposable}, i.e., those that have only trivial direct sum decompositions, play an important role in persistence theory.
A rich family of indecomposable persistence modules is given by \emph{interval modules}, which for an interval $I \subseteq \R$ are defined by
\begin{equation} \label{e:interval module}
    C(I)_t =
    \begin{cases}
        \mathbb{F} & \text{if } t \in I, \\
        0          & \text{otherwise},
    \end{cases}
    \qquad
    \qquad
    C(I)_{s, t} =
    \begin{cases}
        \operatorname{id}_{\mathbb{F}} & \text{if } s, t \in I,\\
        0 & \text{otherwise}.
    \end{cases}    
\end{equation}
These indecomposable interval modules can be used as building blocks for \emph{barcode modules}, which are direct sums of interval modules.
% \[
% \bigoplus_{\lambda \in \Lambda} C(I_{\lambda}).
% \]
The multiset of intervals $\{I_{\lambda}\}_{\lambda \in \Lambda}$ associated to a barcode module is known as its \textit{barcode}. By a version of the Krull--Remak--Schmidt--Azumaya Theorem \cite{MR37832} (see also \cite[Theorem 2.7]{Chazal.2016a} for a specialization to barcode modules), two isomorphic barcode modules have the same barcodes up to a choice of the index set $\Lambda$. Thus, if a persistence module $M$ is isomorphic to a barcode module, the associated barcode is a complete invariant of $M$, and the (non-unique) isomorphism to the barcode module is called a \emph{barcode decomposition}.

We want to use the passage from persistence modules to invariants obtained from decompositions as the last step in our pipeline, so we have to understand which persistence modules admit barcode decompositions.
The most commonly used existence result for barcode decompositions is due to Crawley-Boevey's theorem \cite{Crawley-Boevey.2015}. It guarantees the existence of a barcode decomposition for any \emph{pointwise finite dimensional ($\PFD$)} persistence module, which is a persistence modules $M$ such that $M_t$ is a finite dimensional vector space for all $t \in \R$.

Unfortunately, the $\PFD$ condition is too restrictive for our purposes. In particular, it is not necessarily satisfied in the historical setup of Morse and Tompkins' work on minimal surfaces. In this context, the persistence modules $M$ only have the slightly weaker property of being \emph{q-tame}, which is defined by the rank of the maps $M_{s,t} \colon M_s \to M_t$ being finite for all $s < t$  \cite{Chazal.2016a}. 
Not every q-tame persistence module admits a barcode decomposition in the above sense, as exemplified by the infinite product of interval modules $\prod_{n \in \N_{> 0}} C([0,1/n))$. Yet, there are multiple ways of working around this to still get discrete invariants in the spirit of barcodes \cite{Chazal.2016a, Chazal.2016b, schmahl2020structure}. We briefly recall the approach from \citet{Chazal.2016b}.

@@

The most general stability result is shown by considering as an intermediate step the \emph{interleaving distance} for filtrations and persistence modules \cite{MR2279866}, which measures how far two $\R$-indexed diagrams $M,N$ are from being isomorphic.
A $\delta$-interleaving consists of a pair of natural transformations $(M_t \to N_{t+\delta})_t, (N_t \to M_{t+\delta})_t$ from one diagram to a shifted version of the other and vice versa, which compose to the internal structure maps $(M_t \to M_{t+2\delta})_t, (N_t \to N_{t+2\delta})_t$. Clearly, the case $\delta=0$ describes an isomorphism, and the infimum $\delta$ admitting a $\delta$-interleaving is the interleaving distance between the two diagrams.
As it turns out, the stability of persistence barcodes can then be described as a sequence of 1-Lipschitz transformations: from functions (with the supremum norm) to filtrations by sublevel sets (with the interleaving distance on diagrams of topological spaces), to persistent homology (with interleaving distance on persistence modules), and finally to persistence diagrams (with the bottleneck distance).
In this approach, by far the most difficult step is showing that passing from persistence modules to persistence diagrams is stable, a result which is known as the Algebraic Stability Theorem  \cite{10.1145/1542362.1542407,Chazal.2016a,MR3333456}.

>>>>

By definition, applying to a filtered space $\{X_t\}_{t \in \R}$ a functor from $\Top$ to $\Vect$ yields for every $t \in \R$ a vector space~$M_t$ and for any pair~$s, t \in \R$ with~$s \leq t$ a linear map~$M_{s,t} \colon M_s \to M_t$ such that $M_{t,t}$ is the identity and the composition $M_{s,t} \circ M_{r,s}$ is equal to $M_{r,t}$ for any triple $r \leq s \leq t$.
In other words, a functor from the real numbers considered as a category via its poset structure to the category of vector spaces.
Such functors are referred to as \emph{persistence modules}.
A morphism of persistence modules $\varphi \colon M \to N$ is a natural transformation, i.e., an assignment of a linear map $\varphi_t \colon M_t \to N_t$ for every $t \in \R$ making the diagram
\begin{equation*}
    \begin{tikzcd}
    M_{s} \arrow[r, "M_{s,t}"] \arrow[d] & M_{t} \arrow[d] \\
    N_{s} \arrow[r, "N_{s,t}"] & N_{t}
    \end{tikzcd}
\end{equation*}
commute for all pairs $s \leq t$.

Usual constructions that work in the category of vector spaces over the field $\F$ can be transferred to the category of persistence modules by applying them pointwise.
For example, the kernel and cokernel of a morphism, as well as the direct sum of persistence modules are well-defined.
Persistence modules that are \emph{indecomposable}, i.e., those that have only trivial direct sum decompositions, play an important role in persistence theory.
A rich family of indecomposable persistence modules is given by \emph{interval modules}, which for an interval $I \subseteq \R$ are defined by
\begin{equation} \label{e:interval module}
    C(I)_t =
    \begin{cases}
        \mathbb{F} & \text{if } t \in I, \\
        0          & \text{otherwise},
    \end{cases}
    \qquad
    \qquad
    C(I)_{s, t} =
    \begin{cases}
        \operatorname{id}_{\mathbb{F}} & \text{if } s, t \in I,\\
        0 & \text{otherwise}.
    \end{cases}    
\end{equation}
These indecomposable interval modules can be used as building blocks for \emph{barcode modules}, which are direct sums of interval modules.
Given a barcode module $\bigoplus_{\lambda \in \Lambda} C(I_{\lambda})$, the associated multiset of intervals $\{I_{\lambda}\}_{\lambda \in \Lambda}$ is known as its \textit{barcode}.
By a version of the Krull--Remak--Schmidt--Azumaya Theorem \cite{MR37832} (see also \cite[Theorem 2.7]{Chazal.2016a} for a specialization to barcode modules), two isomorphic barcode modules have the same barcodes up to a choice of the index set $\Lambda$.
Thus, if a persistence module $M$ is provided with an isomorphism to a barcode module, referred to as a \emph{barcode decomposition}, the associated barcode is a complete invariant of $M$.

Since the passage from persistence modules to these discrete invariants is the last step in the persistence pipeline, understanding which persistence modules admit barcode decompositions is key.
The most commonly used existence result for barcode decompositions is due to Crawley-Boevey's theorem \cite{Crawley-Boevey.2015}.
It guarantees the existence of a barcode decomposition for any \emph{pointwise finite dimensional ($\PFD$)} persistence module, which is a persistence modules $M$ such that $M_t$ is a finite dimensional vector space for all $t \in \R$.

Unfortunately, the $\PFD$ condition is too restrictive for many purposes.
In particular, it is unsuited for the the applications of Morse and Tompkins in minimal surfaces theory.
An appropriate weakening of $\PFD$ for more general settings is the notion of \emph{q-tame}.
A persistence module $M$ is q-tame if the rank of the map $M_{s,t} \colon M_s \to M_t$ is finite dimensional for all $s < t$  \cite{Chazal.2016a}.
As exemplified by the infinite product of interval modules $\prod_{n \in \N_{> 0}} C([0,1/n))$, not every q-tame persistence module admits a barcode decomposition in the above sense.
Yet, there are multiple ways of working around this to still get discrete invariants in the spirit of barcodes \cite{Chazal.2016a, Chazal.2016b, schmahl2020structure}.
We briefly recall the approach from \citet{Chazal.2016b}.

@@

The most general stability result is shown by considering as an intermediate step the metric space of filtrations and persistence modules equipped with the \emph{interleaving distance} \cite{MR2279866}.
Given two functors $M,N \colon \R \to \mathsf C$, where $\mathsf C$ is typically $\Top$ or $\Vect$, a $\delta$-interleaving consists of a pair of natural transformations $(M_t \to N_{t+\delta})_t$ and $(N_t \to M_{t+\delta})_t$ from one diagram to a shifted version of the other and vice versa, which compose to the structure maps $(M_t \to M_{t+2\delta})_t$ and $(N_t \to N_{t+2\delta})_t$.
Clearly, the case $\delta=0$ describes an isomorphism, and the infimum $\delta$ for which $M$ and $N$ admit a $\delta$-interleaving is defined as the interleaving distance between them.
As it turns out, the stability of persistence barcodes can then be described as a sequence of 1-Lipschitz transformations: from functions (with the supremum norm) to filtrations by sublevel sets (with the interleaving distance on diagrams of topological spaces), to persistent homology (with interleaving distance on persistence modules), and finally to persistence diagrams (with the bottleneck distance).
In this approach, by far the most difficult step is showing that passing from persistence modules to persistence diagrams is stable, a result which is known as the Algebraic Stability Theorem  \cite{10.1145/1542362.1542407,Chazal.2016a,MR3333456}.
